Neon: Nuclear Norm to Beat Muon

### 2025-08-16 conf_article/neon_old.tex
New abstract and preparation for an article

\begin{abstract}
In this article, we develop a new algorithm for optimizing functions of weight matrices that arise in training large language models. By replacing the spectral-norm regularization used in Muon with a nuclear-norm formulation, we obtain a new optimization problem for the update matrix; its solution defines a novel algorithm we call Neon. We derive the optimal solution to this problem and propose a numerically efficient method to compute it via the Lanczos algorithm. We analyze Neon from a theoretical perspective and provide an extensive empirical study across diverse relevant tasks and settings.

New variant:
\iffalse This article investigates utilization of matrix norms besides the spectral norm in linear optimization oracle (LMO) framework. Making use of the nuclear norm, the mean between the nuclear norm and the Frobenius norm we name the F*-norm, and the dual to F*-norm, we develop 3 new algorithms for optimizing functions of weight matrices that arise in training large language models.\fi

In this article, we develop a family of 3 new algorithms for optimizing functions of weight matrices that arise in training large language models. By replacing the spectral norm, which is used when deriving Muon within the linear optimization oracle framework (LMO), with the nuclear norm, the mean between the nuclear norm and the Frobenius norm we name the F*-norm, and the dual to F*-norm we name F2-norm, we obtain 3 new optimization problems for the update matrix; their solutions define 3 novel algorithms we name Neon, F*-Neon, and F2-Muon respectively. We derive the optimal solutions to these problems and propose a numerically efficient method to compute updates for Neon and F*-Neon via the Lanczos algorithm (while an update of F2-Muon turns out to be a mean between the Muon update and the Normalized SGD update). We analyze all 3 algorithms from a theoretical perspective and provide an extensive empirical study across diverse relevant tasks and settings.

\section{Idea}
\iffalse
The goal of the project is to make variations on Muon to speed it up. Recently, authors of \cite{bernstein2024oldoptimizernewnorm} have proposed to derive update step for an optimizer as the solution of the certain optimization problem. This approach can be utilized to derive Muon \cite{jordan2024muon}, a novel algorithm for fast training of neural networks. Instead of using spectral norm, used to derive Muon, we utilize nuclear norm to produce a new optimization algorithm.
\fi
\subsection{Problem (Project description)}

@@

% Removed template instructions sections
\section{Related Work}
Our review primarily focuses on the Muon and Shampoo optimizers, as our algorithm extends the ideas used to derive these methods. We highlight the advantages and disadvantages of these approaches, the unique effects they introduce, and compare them to Neon.

\subsection{Muon optimizer}

In the previous section, we described the theoretical foundation behind the weight update step in the Muon optimizer, but we did not discuss how to obtain the required matrices in practice. The update step is defined by \eqref{eqn:update_muon}, which requires $UV^T$ with $U$ and $V^T$ from the singular value decomposition of the gradient-like matrix $G$. A naive solution would involve computing the SVD of $G$ and constructing the required expression. However, the developers of the Muon optimizer introduced a workaround using Newton-Schulz iterations \cite{jordan2024muon}. The Newton-Schulz iterations from the original article \cite{jordan2024muon} require 10 matrix-matrix multiplications to achieve the desired accuracy. The asymptotic complexity of such an operation is identical to that of SVD and equals $O(mn \min\{m, n\})$ for an $m \times n$ matrix. Nevertheless, matrix multiplication on modern GPUs can be performed much more efficiently.

The performance of Muon in training large language models was tested \cite{liu2025muon} against AdamW. The testing demonstrated excellent performance by Muon, which was approximately 2 times more efficient in terms of FLOPs required to reach a certain loss value. This is even more remarkable when considering the cost of one iteration: Muon requires an additional $O(mn \min\{m, n\})$ FLOPs per $m \times n$ matrix, while AdamW needs only $O(mn)$.

Another interesting discovery about the Muon optimizer is that it accelerates grokking \cite{tveit2025muonoptimizeracceleratesgrokking}. In the test problem, Muon achieved grokking significantly faster than AdamW in terms of passed epochs, with a mean grokking epoch of 102.89 for Muon and 153.09 for AdamW. The authors suggest that this may be due to the fact that Muon stimulates broader exploration by orthogonalizing the gradient matrix, thus avoiding memorization.

Recently, theoretical guarantees for Muon convergence have been derived \cite{li2025noteconvergencemuon}. In particular, in the $L$-smooth convex case, it achieves $O(1/T^{\frac{1}{2}})$ (with full gradient) and $O(1/T^{\frac{1}{4}})$ (with stochastic gradient) bounds on the Frobenius norm of the gradient or the mathematical expectation of the gradient norm, respectively, where $T$ is the number of iterations.

\subsection{Shampoo optimizer}

Another optimizer that exploits the matrix (and even tensor) structure of weights in neural networks is the Shampoo optimizer. We avoid a detailed description of this method here and refer the reader to the original article \cite{gupta2018shampoopreconditionedstochastictensor}, but we outline the key properties of Shampoo and its relation to the Muon optimizer.

The Shampoo optimizer uses left and right preconditioning for the gradient-like matrix, leveling its spectrum. The preconditioners are computed from the exponentially averaged gradients, and their computation requires $O(n^3 + m^3)$ per $m \times n$ matrix. This exponential averaging is a key feature that provides several distinct interpretations for the preconditioners. They can be viewed as an approximation of the Gauss-Newton component of the Hessian or the first step of the power iteration algorithm for computing the optimal Kronecker product approximation \cite{morwani2025a}. With exponential averaging turned off, the update step of Shampoo can be simplified and becomes identical to that of Muon \cite{jordan2024muon}.

Convergence analysis is presented in the original article \cite{gupta2018shampoopreconditionedstochastictensor}. Shampoo achieves $O(1/T^{\frac{1}{2}})$ convergence for the loss function value in the $L$-smooth convex case, where $T$ is the number of iterations.

\subsection{Synthesis and Neon's position}

Recent developments in optimization techniques show that utilizing the matrix structure of weights in neural networks can be very beneficial. Optimizers following this path converge faster in terms of iterations or epochs and often even FLOPs, but have a high iteration cost. Neon seeks to decrease the iteration cost while preserving fast convergence.

While the uniqie advantages and effects introduced by Neon are yet to be discovered, we can already say that our new optimizer introduces additional overhead of $O(mn)$ FLOPS on average per $m \times n$ weight matrix, which is much better then $O(m n \min\{m, n\})$ for Muon optimizer and $O(n^3 + m^3)$ for Shampoo.

\subsection{Optimization Strategies for Efficient Large Language Model Training}
The unprecedented scale of modern Large Language Models (LLMs) has pushed traditional optimizers like AdamW \cite{Loshchilov2017FixingWD} to their limits in terms of computational efficiency and convergence speed \cite{liu2025muon,chen2025cosmoshybridadaptive}. This challenge has catalyzed research into more sophisticated optimization approaches that can maintain or improve performance while reducing training costs.

The Muon optimizer  has emerged as a promising alternative based on matrix orthogonalization principles. Scaling Muon to billion-parameter LLMs required two key adaptations: the integration of L2 weight decay for stability and the implementation of per-parameter update scaling to handle diverse parameter distributions efficiently . Empirical evaluations demonstrate that Muon can match or exceed AdamW's model quality while requiring only about 52 of the training FLOPs. The successful training of the Moonlight model series, including a 16B-parameter Mixture-of-Experts model, validates Muon's practicality for production-scale applications \cite{liu2025muon}.

Building on this foundation, hybrid approaches like COSMOS \cite{chen2025cosmoshybridadaptive} further enhance efficiency by combining optimization techniques based on gradient structure. COSMOS applies computationally intensive updates to a low-dimensional "leading eigensubspace" while using memory-efficient methods like Muon for the remaining parameters. This approach maintains convergence benefits while substantially reducing memory requirements. For distributed training environments, optimizers like Dion \cite{ahn2025dioncommunicationefficientoptimizerlarge} specifically target communication efficiency by minimizing data exchange between workers through distributed orthonormalization techniques.

>>>>

\begin{abstract}
In this paper, we develop a new algorithm for optimization of functions of weight matrices, which are typical for training large language models. Changing spectral norm, which was used to derive Muon, to nuclear norm, we pose a new optimization problem for an update matrix, solution of which defines a novel algorithm we name Neon. To make it feasible, we use Lanczos algorithm to find a required step. After providing theoretical guarantees of Neon convergence, we compare performances of Neon, Muon, and Adam on training multilayer perceptron, convolutional neural network and NanoGPT.
\end{abstract}

\section{Idea}
The goal of the project is to make variations on Muon to speed it up. Recently, authors of \cite{bernstein2024oldoptimizernewnorm} have proposed to derive update step for an optimizer as the solution of the certain optimization problem. This approach can be utilized to derive Muon \cite{jordan2024muon}, a novel algorithm for fast training of neural networks. Instead of using spectral norm, used to derive Muon, we utilize nuclear norm to produce a new optimization algorithm.

\subsection{Problem (Project description)}

@@

% Removed template instructions sections

### 2025-08-16 old_report/muon.tex
New abstract and preparation for an article

\documentclass{article} % For LaTeX2e
\usepackage{icomp2024_conference,times}

% Optional math command from https://github.com/goodfeli/dlbook_notation.
\input{math_commands.tex}

\usepackage{hyperref}
\usepackage{url}
\usepackage{amsmath,amssymb}
\usepackage{amsthm}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{subcaption}
\newtheorem{lemma}{Lemma}
\newcommand{\norm}[1]{\lVert #1\rVert}

\title{From Muon to Neon: Introducing Nuclear Norm to Large Matrices}

% Authors must not appear in the submitted version. They should be hidden
% as long as the \icompfinalcopy macro remains commented out below.
% Non-anonymous submissions will be rejected without review.

\author{Alexey Kravatskiy \\
Moscow Institute of Physics and Technology (MIPT) \\
\texttt{kravtskii.aiu@phystech.edu} \\
\And
Ivan Kozyrev \\
Moscow Institute of Physics and Technology (MIPT) \\
\texttt{kozyrev.in@phystech.edu} \\
\And
Nikolai Kozlov \\
Moscow Institute of Physics and Technology (MIPT) \\
\texttt{kozlov.na@phystech.edu} \\
\And
Alexander Vinogradov \\
Moscow Institute of Physics and Technology (MIPT) \\
\texttt{vinogradov.am@phystech.edu} \\
\And
Daniil Merkulov \\
Moscow Institute of Physics and Technology (MIPT), Skoltech, HSE, Sber \\
\texttt{daniil.merkulov@phystech.edu} \\
\And
Ivan Oseledets \\
AIRI, Skoltech \\
\texttt{i.oseledets@skoltech.ru}
}

% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to \LaTeX{} to determine where to break
% the lines. Using \AND forces a linebreak at that point. So, if \LaTeX{}
% puts 3 of 4 authors names on the first line, and the last on the second
% line, try using \AND instead of \And before the third author name.

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}

%\icompfinalcopy % Uncomment for camera-ready version, but NOT for submission.
\begin{document}

@@

\begin{abstract}
In this article, we develop a new algorithm for optimizing functions of weight matrices that arise in training large language models. By replacing the spectral-norm regularization used in Muon with a nuclear-norm formulation, we obtain a new optimization problem for the update matrix; its solution defines a novel algorithm we call Neon. We derive the optimal solution to this problem and propose a numerically efficient method to compute it via the Lanczos algorithm. We analyze Neon from a theoretical perspective and provide an extensive empirical study across diverse relevant tasks and settings.

New variant:
\iffalse This article investigates utilization of matrix norms besides the spectral norm in linear optimization oracle (LMO) framework. Making use of the nuclear norm, the mean between the nuclear norm and the Frobenius norm we name the F*-norm, and the dual to F*-norm, we develop 3 new algorithms for optimizing functions of weight matrices that arise in training large language models.\fi

In this article, we develop a family of 3 new algorithms for optimizing functions of weight matrices that arise in training large language models. By replacing the spectral norm, which is used when deriving Muon within the linear optimization oracle framework (LMO), with the nuclear norm, the mean between the nuclear norm and the Frobenius norm we name the F*-norm, and the dual to F*-norm we name F2-norm, we obtain 3 new optimization problems for the update matrix; their solutions define 3 novel algorithms we name Neon, F*-Neon, and F2-Muon respectively. We derive the optimal solutions to these problems and propose a numerically efficient method to compute updates for Neon and F*-Neon via the Lanczos algorithm (while an update of F2-Muon turns out to be a mean between the Muon update and the Normalized SGD update). We analyze all 3 algorithms from a theoretical perspective and provide an extensive empirical study across diverse relevant tasks and settings.

\section{Idea}
\iffalse
The goal of the project is to make variations on Muon to speed it up. Recently, authors of \cite{bernstein2024oldoptimizernewnorm} have proposed to derive update step for an optimizer as the solution of the certain optimization problem. This approach can be utilized to derive Muon \cite{jordan2024muon}, a novel algorithm for fast training of neural networks. Instead of using spectral norm, used to derive Muon, we utilize nuclear norm to produce a new optimization algorithm.
\fi
\subsection{Problem (Project description)}

@@

% Removed template instructions sections
\section{Related Work}
Our review primarily focuses on the Muon and Shampoo optimizers, as our algorithm extends the ideas used to derive these methods. We highlight the advantages and disadvantages of these approaches, the unique effects they introduce, and compare them to Neon.

@@

\begin{proof}
Let us denote $r = \min\{m, n\}$. Then by Von Neumann's trace inequality,
\begin{equation*}
    \vert\langle G, \delta W \rangle\vert \leq \sum_{i=1}^r \sigma_i(G)\sigma_i(\delta W)
    \Rightarrow
    \langle G, \delta W \rangle \geq -\sum_{i=1}^r \sigma_i(G)\sigma_i(\delta W)\,.
\end{equation*}
Thus, expressing nuclear norm through singular values, we can write down
\begin{equation*}
    f(\delta W) \geq -\sum_{i=1}^r \sigma_i(G)\sigma_i(\delta W) + \lambda\left(\sum_{i=1}^{r}\sigma_i(\delta W)\right)^2 \geq \min_{d_1,\dots,d_r \geq 0}-\sum_{i=1}^r \sigma_i(G)d_i + \lambda\left(\sum_{i=1}^{r}d_i\right)^2.
\end{equation*}
By Karush-Kuhn-Tucker theorem, necessary conditions of minimum are
\begin{equation*}
    \left(d_i \geq 0\ \text{and}\ -\sigma_i(G) + 2\lambda\sum_{j=1}^{r}d_j = 0\right)\ \text{or}\ \left(d_i = 0\ \text{and}\ -\sigma_i(G) + 2\lambda\sum_{j=1}^{r}d_j \geq 0\right) \,\, i = 1,\dots,r\,.
\end{equation*}
These conditions simplify to 
\begin{equation*}
    \sum_{i \in S}d_i = \sigma_1(G)\,,\ \begin{cases}
        d_i \geq 0 & \text{if}\ \sigma_i(G) = \sigma_1(G)\,, \\
        d_i = 0 & \text{otherwise}\,.
    \end{cases}
\end{equation*}
All points satisfying those conditions deliver minimum, and
\begin{equation*}
    f(\delta W) \geq -\frac{\sigma_1^2(G)}{4\lambda}\,.
\end{equation*}

@@

\begin{proof}
Analogously to the proof of Lemma~\ref{lem:opt_star}, we can use Von Neumann's trace inequality to write down
\begin{equation*}
    f(\delta W) \geq -\sum_{i=1}^r \sigma_i(G)\sigma_i(\delta W) + \frac{\lambda}{2}\left(\sum_{i=1}^{r}\sigma_i(\delta W)\right)^2 + \frac{\lambda}{2}\sum_{i=1}^{r}\sigma_i^2(\delta W)\,,
\end{equation*}
\begin{equation}\label{eqn:f_min_F_star}
    f(\delta W) \geq \frac{1}{\lambda}\min_{d_1,\dots,d_r \geq 0}-\sum_{i=1}^r \sigma_i(G)d_i + \frac{1}{2}\left(\sum_{i=1}^{r}d_i\right)^2 + \frac{1}{2}\left(\sum_{i=1}^{r}d_i^2\right)\,,
\end{equation}
By Karush-Kuhn-Tucker theorem, neccesary conditions of minimum are
\begin{equation*}
    \left(d_i \geq 0\ \text{and}\ -\sigma_i(G) + \sum_{j=1}^{r}d_j +  d_i = 0\right)\ \text{or}\ \left(d_i = 0\ \text{and}\ -\sigma_i(G) + \sum_{j=1}^{r}d_j + d_i \geq 0\right) \,\, i = 1,\dots,r\,.
\end{equation*}
Denoting $\tau = \sum_{i = 1}^r d_i$ gives $d_i = [\sigma_i(G) - \tau]_+$, where $\tau$ satisfies
\begin{equation}
    \sum_{i=1}^n [\sigma_i(G) - \tau]_+ = \tau\,.
\end{equation}
Inserting found minimum point into~\eqref{eqn:f_min_F_star} yields
\begin{equation*}
    f(\delta W) \geq -\sum_{i = 1}^r d_i(\tau + d_i) + \frac{\tau^2}{2\lambda} + \frac{1}{2\lambda}\sum_{i=1}^{r}d_i^2 = -\frac{1}{2 \lambda} \left( \tau^2 + \sum_{i=1}^r d_i^2 \right)\,.
\end{equation*}

@@

\section{Prototyping phase report}
\begin{enumerate}
    \item Update rule is derived, see idea
    \item Update rule methods are tested: power iteration vs Lanczos (see \ref{tab:matrix_methods})
    \item Recorded the distribution of singular values of gradients during NanoGPT training (see Firgures ~\ref{fig:svd_all}, \ref{fig:svd_all}).
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{figs/mlp24/svd_evolution_all.png}}
        \caption{Singular values of $50257\times 1280$ layer via 200 iterations}
        \label{fig:svd_all}
    \end{figure}
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{figs/mlp24/svd_evolution.png}}
        \caption{Singular values of  of $50257\times 1280$ layer for 5-th,45-th, 65-th,175-th and 200-th iteration}
        \label{fig:svd}
    \end{figure}

    \item NanoGPT is tested on Muon and Adam. For now, Neon (rank-1 version) does not converge (see Figures ~\ref{fig:train_loss} \ref{fig:val_loss})
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{figs/mlp24/loss_train.png}}
        \caption{Train loss}
        \label{fig:train_loss}
    \end{figure}
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{figs/mlp24/loss_val.png}}
        \caption{Validation loss}
        \label{fig:val_loss}
    \end{figure}

@@

    \item Neon (rank-1 version), Muon, AdamW and SGD are compared on MLP and CNN (see Figures~\ref{fig:mlp_epochs}, \ref{fig:cnn_epochs}, \ref{fig:mlp_time}, and~\ref{fig:cnn_time}). All methods work correctly, but again there is the problem with which one is the fastest (for now, it's SGD).
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{figs/mlp0605/loss_vs_epochs_mlp.pdf}}
        \caption{MLP: self.linear1 = nn.Linear(32*32*3, 512), self.linear2 = nn.Linear(512, 10), self.activ = nn.GELU()}
        \label{fig:mlp_epochs}
    \end{figure}
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{figs/mlp24/accuracy_vs_epochs_moderate.png}}
        \caption{CNN: 2 convolutional blocks, 2 fully connected layers, activation + dropout}
        \label{fig:cnn_epochs}
    \end{figure}
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{figs/mlp0605/accuracy_vs_time_mlp.pdf}}
        \caption{MLP: wallclock time measurements}
        \label{fig:mlp_time}
    \end{figure}
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{figs/mlp24/accuracy_vs_time_moderate.png}}
        \caption{CNN: wallclock time measurements}
        \label{fig:cnn_time}
    \end{figure}
    \begin{table}[h!]
        \centering
        \begin{tabular}{c|c|c|c}
        \hline
        Method                              & rtol        & k & time,s \\
        \hline 
        Power Iterations                    & 0.01        & 1 & 7.7\\ 
        SVDS (thick-restart Lanczos method) & 0.01        & 1 & 0.18\\
        PCA Low Rank (RSVD)                 & 0.01        & 1 & 1.15\\
        SVDS (thick-restart Lanczos method) & 0.01        & 10 & 0.47\\
        PCA Low Rank (RSVD)                 & 0.01        & 10 & 19.4\\
        SVDS (thick-restart Lanczos method) & 0.01        & 100 & 1.96\\
        PCA Low Rank (RSVD)                 & 0.01        & 100 & 170\\
        \end{tabular}
        \caption{k-rank updated comparison}
        Comparison of different numerical methods to calculate k-rank update on $5000\times5000$ matrix of real numbers, rtol is an error in Frobenius norm relative to the k-rank approximation of truncated svd. During the research it was noted that rsvd can give good and fast approximation for singular values, but the matrix of approximation is far from the one given by truncated svd, while Lanczos method gives good and fast approximation for a matrix, but not so good approximation for singular values. 
        \label{tab:matrix_methods}
    \end{table}
\end{enumerate}

\bibliography{icomp2024_conference}
\bibliographystyle{icomp2024_conference}

\appendix
\section{Appendix}
You may include other additional sections here.

>>>>

%% ============================================
%% ================ Preambule =================
%% ============================================
\documentclass[]{scrartcl}
\usepackage[margin = 0.5in]{geometry}

\usepackage[pdftex,unicode, 
colorlinks=true,
linkcolor = blue]{hyperref}	% нумерование страниц, ссылки!!!!ИМЕННО В ТАКОМ ПОРЯДКЕ СО СЛЕДУЮЩИМ ПАКЕТОМ
%\usepackage[warn]{mathtext}				% Поддержка русского текста в формулах
\usepackage[T1, T2A]{fontenc}			% Пакет выбора кодировки и шрифтов
\usepackage[utf8]{inputenc} 			% любая желаемая кодировка
\usepackage[english]{babel}		% поддержка русского языка
\usepackage{wrapfig}					% Плавающие картинки
\usepackage{amssymb, amsmath}			% стилевой пакет для формул
\usepackage{amsthm}                     % поддержка теорем
\usepackage{algorithm}
\usepackage{algorithmic} 
\usepackage[backend=biber]{biblatex}
\addbibresource{muon.bib}

\ifpdf
\usepackage{cmap} 				% чтобы работал поиск по PDF
\usepackage[pdftex]{graphicx}
%\usepackage{pgfplotstable}		% Для вставки таблиц.
\pdfcompresslevel=9 			% сжимать PDF
\else
\usepackage{graphicx}
\fi

\graphicspath{{./figures/}}
\usepackage{subcaption}

\newtheorem{lemma}{Lemma}

%custom commands
\newcommand{\norm}[1]{\Vert{#1}\Vert}

%% ============================================
%% ================ Info =================
%% ============================================
\title{Neon: Nuclear Norm to Beat Muon}

\author{
  Alexey Kravatskiy\\
  \texttt{kravtskii.aiu@phystech.edu}
  \and
  Ivan Kozyrev\\
  \texttt{kozyrev.in@phystech.edu}
  \and
  Nikolay Kozlov \\
  \texttt{kozlov.na@phystech.edu}
  \and
  Alexander Vinogradov \\
  \texttt{vinogradov.am@phystech.edu}
  \and
  Daniil Merkulov \\
  \texttt{daniil.merkulov@phystech.edu}
  \and
  Ivan Oseledets \\
  \texttt{i.oseledets@scholtech.ru}
}

%\author{\begin{tabular}{c c}
%	  	 Alexey Kravatsky & Nikolay Kozlov & Ivan Kozyrev \\
%		 \texttt{kravatskii.aiu@phystech.edu} & \texttt{kozlov.n?@phystech.edu} & \texttt{kozyrev.i?@phystech.edu}  
%		\end{tabular}}
\date{\today}

\begin{document}

@@

\begin{abstract}
In this paper, we develop a new algorithm for optimization of functions of weight matrices, which are typical for training large language models. Changing spectral norm, which was used to derive Muon, to nuclear norm, we pose a new optimization problem for an update matrix, solution of which defines a novel algorithm we name Neon. To make it feasible, we use Lanczos algorithm to find a required step. After providing theoretical guarantees of Neon convergence, we compare performances of Neon, Muon, and Adam on training multilayer perceptron, convolutional neural network and NanoGPT.

\end{abstract}

\section{Idea}
The goal of the project is to make variations on Muon to speed it up. Recently, authors of \cite{bernstein2024oldoptimizernewnorm} have proposed to derive update step for an optimizer as the solution of the certain optimization problem. This approach can be utilized to derive Muon \cite{jordan2024muon}, a novel algorithm for fast training of neural networks. Instead of using spectral norm, used to derive Muon, we utilize nuclear norm to produce a new optimization algorithm. %Then, we add momentum to update. After finalizing the algorithm, we test it on MLP, and, if results are satisfactory, on transformers (probably something from huggingface). The result will be a fast algorithm, which we will convert into a new optimizer class for PyTorch, as was done with Muon.%

\subsection{Problem (Project description)}

@@

\section{Outcomes}
The main results of the project are expected to be:
\begin{enumerate}
    \item Deriving of Neon and argumentation of the numerical methods which will be used in the algorithm
    \item Code implementation of Neon as PyTorch optimizer class
    \item Numerical experiments comparing performance of training with Neon, Muon, SVD and Adam on several architectures: MLP, CNN, and NanoGPT.
    \item Participation in CIFAR-10 challenge with Neon (if the alrothim turns out to be competitive)
    \item Article describing all the results. If they are satisfactory, we intend to be in time to apply for NeurIPS 2025. It is a challenge, but I think we must give it a try.
\end{enumerate}

\section{Literature review}
Our review primarily focuses on the Muon and Shampoo optimizers, as our algorithm extends the ideas used to derive these methods. We highlight the advantages and disadvantages of these approaches, the unique effects they introduce, and compare them to Neon.

@@

\begin{proof}
Let us denote $r = \min\{m, n\}$. Then by Von Neumann's trace inequality,
\begin{equation*}
    \vert\langle G, \delta W \rangle\vert \leq \sum_{i=1}^r \sigma_i(G)\sigma_i(\delta W)
    \Rightarrow
    \langle G, \delta W \rangle \geq -\sum_{i=1}^r \sigma_i(G)\sigma_i(\delta W)\,.
\end{equation*}
Thus, expressing nuclear norm through singular values, we can write down
\begin{equation*}
    f(\delta W) \geq -\sum_{i=1}^r \sigma_i(G)\sigma_i(\delta W) + \lambda\left(\sum_{i=1}^{r}\sigma_i(\delta W)\right)^2 \geq \min_{d_1,\dots,d_r \geq 0}-\sum_{i=1}^r \sigma_i(G)d_i + \lambda\left(\sum_{i=1}^{r}d_i\right)^2.
\end{equation*}
By Karush-Kuhn-Tucker theorem, necessary conditions of minimum are
\begin{equation*}
    \left(d_i \geq 0\ \text{and}\ -\sigma_i(G) + 2\lambda\sum_{j=1}^{r}d_j = 0\right)\ \text{or}\ \left(d_i = 0\ \text{and}\ -\sigma_i(G) + 2\lambda\sum_{j=1}^{r}d_j \geq 0\right) \,,\ i = 1,\dots,r\,.
\end{equation*}
These conditions simplify to 
\begin{equation*}
    \sum_{i \in S}d_i = \sigma_1(G)\,,\ \begin{cases}
        d_i \geq 0 & \text{if}\ \sigma_i(G) = \sigma_1(G)\,, \\
        d_i = 0 & \text{otherwise}\,.
    \end{cases}
\end{equation*}
All points satisfying those conditions deliver minimum, and
\begin{equation*}
    f(\delta W) \geq -\frac{\sigma_1^2(G)}{4\lambda}\,.
\end{equation*}

@@

\begin{proof}
Analogously to the proof of Lemma~\ref{lem:opt_star}, we can use Von Neumann's trace inequality to write down
\begin{equation*}
    f(\delta W) \geq -\sum_{i=1}^r \sigma_i(G)\sigma_i(\delta W) + \frac{\lambda}{2}\left(\sum_{i=1}^{r}\sigma_i(\delta W)\right)^2 + \frac{\lambda}{2}\sum_{i=1}^{r}\sigma_i^2(\delta W)\,,
\end{equation*}
\begin{equation}\label{eqn:f_min_F_star}
    f(\delta W) \geq \frac{1}{\lambda}\min_{d_1,\dots,d_r \geq 0}-\sum_{i=1}^r \sigma_i(G)d_i + \frac{1}{2}\left(\sum_{i=1}^{r}d_i\right)^2 + \frac{1}{2}\left(\sum_{i=1}^{r}d_i^2\right)\,.
\end{equation}
By Karush-Kuhn-Tucker theorem, neccesary conditions of minimum are
\begin{equation*}
    \left(d_i \geq 0\ \text{and}\ -\sigma_i(G) + \sum_{j=1}^{r}d_j +  d_i = 0\right)\ \text{or}\ \left(d_i = 0\ \text{and}\ -\sigma_i(G) + \sum_{j=1}^{r}d_j + d_i \geq 0\right) \,,\ i = 1,\dots,r\,.
\end{equation*}
Denoting $\tau = \sum_{i = 1}^r d_i$ gives $d_i = [\sigma_i(G) - \tau]_+$, where $\tau$ satisfies
\begin{equation}
    \sum_{i=1}^n [\sigma_i(G) - \tau]_+ = \tau\,.
\end{equation}
Inserting found minimum point into~\eqref{eqn:f_min_F_star} yields
\begin{equation*}
    f(\delta W) \geq -\sum_{i = 1}^r d_i(\tau + d_i) + \frac{\tau^2}{2\lambda} + \frac{1}{2\lambda}\sum_{i=1}^{r}d_i^2 = -\frac{1}{2 \lambda} \left( \tau^2 + \sum_{i=1}^r d_i^2 \right)\,.
\end{equation*}

@@

\section{Prototyping phase report}
\begin{enumerate}
    \item Update rule is derived, see idea
    \item Update rule methods are tested: power iteration vs Lanczos (see \ref{tab:matrix_methods})
    \item Recorded the distribution of singular values of gradients during NanoGPT training (see Firgures ~\ref{fig:svd_all}, \ref{fig:svd_all}).
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{../figs/mlp24/svd_evolution_all.png}}
        \caption{Singular values of $50257\times 1280$ layer via 200 iterations}
        \label{fig:svd_all}
    \end{figure}
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{../figs/mlp24/svd_evolution.png}}
        \caption{Singular values of  of $50257\times 1280$ layer for 5-th,45-th, 65-th,175-th and 200-th iteration}
        \label{fig:svd}
    \end{figure}

    \item NanoGPT is tested on Muon and Adam. For now, Neon (rank-1 version) does not converge (see Figures ~\ref{fig:train_loss} \ref{fig:val_loss})
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{../figs/mlp24/loss_train.png}}
        \caption{Train loss}
        \label{fig:train_loss}
    \end{figure}
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{../figs/mlp24/loss_val.png}}
        \caption{Validation loss}
        \label{fig:val_loss}
    \end{figure}

@@

    \item Neon (rank-1 version), Muon, AdamW and SGD are compared on MLP and CNN (see Figures~\ref{fig:mlp_epochs}, \ref{fig:cnn_epochs}, \ref{fig:mlp_time}, and~\ref{fig:cnn_time}). All methods work correctly, but again there is the problem with which one is the fastest (for now, it's SGD).
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{../figs/mlp0605/loss_vs_epochs_mlp.pdf}}
        \caption{MLP: self.linear1 = nn.Linear(32*32*3, 512), self.linear2 = nn.Linear(512, 10), self.activ = nn.GELU()}
        \label{fig:mlp_epochs}
    \end{figure}
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{../figs/mlp24/accuracy_vs_epochs_moderate.png}}
        \caption{CNN: 2 convolutional blocks, 2 fully connected layers, activation + dropout}
        \label{fig:cnn_epochs}
    \end{figure}
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{../figs/mlp0605/accuracy_vs_time_mlp.pdf}}
        \caption{MLP: wallclock time measurements}
        \label{fig:mlp_time}
    \end{figure}
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{../figs/mlp24/accuracy_vs_time_moderate.png}}
        \caption{CNN: wallclock time measurements}
        \label{fig:cnn_time}
    \end{figure}
    \begin{table}[h!]
        \centering
        \begin{tabular}{c|c|c|c}
        \hline
        Method                              & rtol        & k & time,s \\
        \hline 
        Power Iterations                    & 0.01        & 1 & 7.7\\ 
        SVDS (thick-restart Lanczos method) & 0.01        & 1 & 0.18\\
        PCA Low Rank (RSVD)                 & 0.01        & 1 & 1.15\\
        SVDS (thick-restart Lanczos method) & 0.01        & 10 & 0.47\\
        PCA Low Rank (RSVD)                 & 0.01        & 10 & 19.4\\
        SVDS (thick-restart Lanczos method) & 0.01        & 100 & 1.96\\
        PCA Low Rank (RSVD)                 & 0.01        & 100 & 170\\
        \end{tabular}
        \caption{k-rank updated comparison}
        Comparison of different numerical methods to calculate k-rank update on $5000\times5000$ matrix of real numbers, rtol is an error in Frobenius norm relative to the k-rank approximation of truncated svd. During the research it was noted that rsvd can give good and fast approximation for singular values, but the matrix of approximation is far from the one given by truncated svd, while Lanczos method gives good and fast approximation for a matrix, but not so good approximation for singular values. 
        \label{tab:matrix_methods}
    \end{table}
\end{enumerate}

\printbibliography
